Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-07-14 Chapter_1/04_Manuscript/Manuscript.tex
Updated manuscript

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding,
colonize new habitats and reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and preservation of dispersal corridors has become a
task of extraordinary importance for conservation authorities worldwide.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their suitability in reality. Individual-based dispersal simulations have
been proposed to address these shortcomings, yet due to the sheer amount of
non-trivial modeling decisions decisions required, a unified and objective
framework to simulate dispersal is missing.

Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful to
investigate \textit{transient} movement behavior and landscape connectivity.

Here, we showcase the use of integrated step selection functions as a simple
individual-based and spatially explicit model to simulate dispersal of the
endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions, thereby parametrizing a fully
mechanistic movement model describing how dispersing wild dogs move through the
landscape. Based on this model, we simulate 80'000 dispersers moving across the
extent of the KAZA-TFCA. and generate a set of maps, each focused on a different
aspect of connectivity. Finally, we discuss the benefits and pitfalls of such a
simulation-based approach and highlight potential improvements to be made in the
future.

@@

% Create linenumbers
\linenumbers

@@

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
path analysis (LCP analysis; \citealp{Adriaensen.2003}) and circuit theory (CT;
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape to infer likely movement corridors. Despite their
intuitive nature and ease of use, both methods make rigorous assumptions about
animal movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be fulfilled by
migrating animals that typically move between a discrete set of habitats through
familiar landscapes. Dispersers, on the other hand, usually move over long
distances into unknown territory and are therefore less likely to be aware of
associated movement costs \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Another
issue of LCPs analysis concerns the fact that least-costly routes, by their very
nature, are only one pixel wide \citep{Pinto.2009}. This neglects the fact that
alternative routes with similar costs may exist and implies that the width of
inferred movement routes depends on the resolution of chosen covariate layers
and may not be biologically meaningful \citep{Diniz.2020}. Although some of
these deficiencies can be addressed using less stringent versions of the LCP
algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness remains.

\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are hardly ever met. For
example, because CT only allows movements from a source cell to its 4 or 8
adjacent cells, it implicitly posits that animals exhibit a perceptual range of
a single pixel. Given that covariate layers are usually resolved with a pixel
size between 30 m x 30 m and 1 km x 1 km, this hardly ever renders the true
capability of animals to perceive the environment. Moreover, CT is built around
the assumption of a complete random walk \citep{Diniz.2020}, entailing that
directional biases cannot be rendered. Nevertheless, directionality is a common
characteristic in animal movement \citep{Bovet.1991, Schultz.2001}, especially
in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}.

\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal \citep{Diniz.2020}. Statements about the expected
duration required to traverse a certain corridor are therefore impossible.
Likewise, because movement is not modeled explicitly, interactions between
movement and habitat preferences of the focal species cannot be rendered.
Connectivity therefore merely arises in result to the landscape structure, which
is usually referred to as structural connectivity. While structural connectivity
yields insights in the \textit{potential} of the landscape to be traversed, it
does not enable to quantify the \textit{actual} gene flow through the area.
Consequently, a functional view on connectivity, which also renders the
behavioral response of the animal with respect to prevailing habitat conditions,
is often more desirable \citep{Tischendorf.2000, Baguette.2013}.

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis. First, GPS
    data and habitat covariates have to be collected. The combined data is then
    analyzed in an integrated step selection model, which enables the
    parametrization of the focal species' habitat and movement kernels. The
    parametrized model is then treated as an individual-based movement model and
    used to simulate dispersal trajectories. Ultimately, simulated trajectories
    serve to produce a set of maps that are pertinent to landscape connectivity.
    This includes a heatmap, indicating the relative traversal frequency across
    the study area, a betweenness map, highlighting movement corridors and
    bottlenecks, and, finally, an inter-patch connectivity map, where the
    frequency of connections and their average duration can be depicted. Photos:
    Whom to cite? Vectronics or Photographers?}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Study Area (90\%)}
The study area was defined by a bounding box centered at -17\degree 13'9''S,
23\degree 56'4''E (\Cref{StudyArea}a) stretching over 1.3 Mio.
km\textsuperscript{2} and encompassing the entire KAZA-TFCA (\Cref{StudyArea}b).
The KAZA-TFCA represents the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. It covers a
total of 520'000 km\textsuperscript{2} and hosts diverse landscapes, ranging
from savanna to grassland and from dry to moist woodland habitats. In its center
lies the Okavango Delta, a dominant hydrogeographical feature and the world's
largest flood-pulsing inland delta. The wet season within the KAZA-TFCA lasts
from November to March and is out of phase with the flood in the Okavango Delta,
which peaks between July and August \citep{McNutt.1996, Wolski.2017}. Although
large portions within the KAZA-TFCA are designated national parks or other
protected areas, considerable human influence remains due to roads, agricultural
sites and settlements and villages that are distributed across the KAZA-TFCA's
landscape.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    encompassing parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green). The dispersal data used in
    this study was collected on a free-ranging African wild dog population
    inhabiting the Moremi National Park in northern Botswana.}
    \label{StudyArea}
  \end{center}
\end{figure}

@@

We ran conditional logistic regression analysis in the r-package {\tt glmmTMB}
to estimate preferences of interest. To handle multiple individuals, we applied
the mixed effects technique developed by \citep{Muff.2020}, which allows to
effectively model random slopes. Thus, we treated animal IDs as random effect
and modeled random slopes for each covariate. We fixed the random intercept
variance at an arbitrary high value of 10\textsuperscript{6} to make use of the
``poission''-trick \citep{Muff.2020}.

@@

\subsection{Convergence (80\%)}
To verify that the number of simulated individuals sufficed to ensure reliable
estimates of connectivity, we evaluated how the relative traversal frequency
across the landscape depended on the number of simulated trajectories.
Specifically, we distributed 1'000 squared ``checkpoints'', each with an extent
of 5 km x 5 km at random locations inside the main study area. We then
determined the relative traversal frequency by simulated trajectories through
each checkpoint for different numbers of simulations (1 to 50'000). We
repeatedly sampled trajectories 100 times and computed the mean traversal
frequency across replicates, as well as the 95\% prediction-interval. We
expected that the mean traversal frequency converges towards a steady state with
increasing simulations.

@@

% On average, step lengths realized by the simulated dispersers (\(\mu_{sl} =
% 2'093\) m, \(\sigma_{sl} = 3'067\)) were slightly shorter than those by observed
% dispersers (\(\mu_{sl} = 2'326\) m, \(\sigma_{sl} = 3'323\)) and simulated
% dispersers moved marginally less directional (\(\mu_{cos(ta)} = 0.057\),
% \(\sigma_{cos(ta)} = 0.071\)) compared to observed dispersers (\(\mu_{cos(ta)} =
% 0.078\), \(\sigma_{cos(ta)} = 0.072\)). These differences in step lengths and
% turning angles can be attributed to minor disparities between habitat conditions
% at the area within which we collected training data and habitat conditions
% within the entire study area.

@@

\subsection{Inter-Patch Connectivity (80\%)}
Results from the analysis of inter-patch connectivity are given in
\Cref{AreasReached}. The figure shows all realized links by simulated dispersers
between national parks and indicates the average duration a disperser had to
move to realize those links. It is again worth pointing out that the figure is
only intended as an example and for clarity we only consider connectivity
between national parks (NPs), albeit plenty of links between other protected
areas exist. As can be seen from the number and thicknesses of arrows,
inter-patch connectivity between NPs in Angola, Namibia, and Botswana is
comparably high. In addition, the bright colors highlight that most dispersal
events between those areas are short. In contrast, we see that connections into
the Kafue NP in Zambia require more steps and are fewer in general. Similarly,
there is a lack of connections into Zimbabwe's Chizarira and Matusadona NP and
the more distant Lower Zambezi and Mana Pools NPs. In some cases, one can also
detect imbalances between ingoing and outgoing links, hinting at potential
source-sink dynamics that occur due to asymmetries in landscape permeability
depending on the origin. For instance, while a large portion of dispersers from
the Chizaria NP in Zimbabwe manage to move into the Hwange NP, there are
comparably few dispersers that succeed in the opposite direction.

@@

\section{Discussion}
\subsection{Short Summary (90\%)}
We used ISSFs to analyse data of dispersing wild dogs and to parametrize a fully
mechanistic movement model describing how dispersers move through the available
landscape. We employed the parametrized model as an individual-based movement
model to simulate 80'000 dispersing wild dogs moving 2'000 steps across the
extent of the KAZA-TFCA, the world's largest transboundary conservation area,
located in southern Africa. Based on simulated dispersal trajectories, we
derived three complementary maps, all geared towards a better understanding of
dispersal and landscape connectivity. These maps included a heatmap, revealing
frequently traversed areas, a betweenness-map, delineating critical dispersal
corridors, and a map of inter-patch connectivity, indicating the presence or
absence of functional links between national parks. We thereby showcase that
ISSFs offer a simple, yet powerful framework to parametrize movement models and
simulate dispersal to assess landscape connectivity. Importantly,
individual-based simulations from ISSFs overcome several conceptual shortcomings
inherent to more traditional connectivity modeling techniques, such as
least-cost path analysis and circuit theory.

\subsection{Movement Model (80 \%)}
Our movement model of dispersing wild dogs consisted of a habitat kernel, a
movement kernel, and their interactions. Together, the kernels in detail
described habitat and movement preferences of dispersing wild dogs, as well as
how movement behavior was affected by habitat conditions. Parameter estimates
from the habitat kernel revealed that dispersing wild dogs avoid water, prefer
its proximity, avoid woodland, prefer shrubs/grassland, and avoid areas
dominated by humans. These findings are consistent with findings from previous
studies on dispersing wild dogs \citep{DaviesMostert.2012, Masenga.2016,
Woodroffe.2019, Oneill.2020}, as well as with a base model that we developed
with a main focus on dispersers' habitat kernel \citep{Hofmann.2021}.

By expanding our base model by a proper movement kernel, we were able to model
several additional complexities inherent to dispersal. For instance, it is well
known that dispersers move with directional persistence and that their step
lengths are correlated with turning angles \citep{Morales.2004, Borger.2012}.
That is, larger steps usually imply smaller turning angles and vice versa. While
this behavior could be captured by sampling turning angles and step lengths from
copula probability distributions \citep{Hodel.2021}, the ISSF framework allowed
us to model similar behavior directly in the movement model. Besides accounting
for directional persistence and correlations between step lengths and turning
angles, we also allowed for interactions that rendered the fact that wild dogs
mainly move during the morning and evening hours, whereas they tend to rest
during the remainder of the day. The final model thus rendered a burst-like
movement behavior, where dispersers realize larger steps during the darker
morning and evening hours, but shorter steps during the rest of the day. While
we accounted for 1. order autocorrelation (i.e. correlation between two
subsequent steps), higher order autocorrelation is conceivable and might be
desirable to model. While ISSFs are conceptually capable of rendering such
behavior, it unfortunately requires vast amounts of GPS data that is not
intercepted by missing fixes, making it impractical to model higher-order
autocorrelation using ISSFs in reality.

By allowing interactions between habitat covariates and movement covariates, we
furthermore accounted for the fact that movement and habitat preferences are
interdependent. For example, the final model retained an interaction between
water cover and step length, showing that dispersers are more likely to realize
short steps (i.e. move slower) in areas covered by water. Likewise, the
parameter estimate for the interaction between water cover and turning angles
revealed that dispersers move less directional when moving across water bodies.
We believe that this behavior is owed to the fact that wild dogs wade or swim
when traversing waterbodies, thus resulting in slower, more tortuous movements.
Besides this, our model also suggested that dipsersers preferably realize
shorter steps when moving through woodland, but larger steps when moving across
shrubs/grassland. This can likely be linked to wild dogs' resting behavior, as
wild dogs usually use open areas to quickly move over long distances
\citep{Abrahms.2017} but seek shade and protection below the woodland canopy
when resting \citep{Creel.2002}. However, we did not consider three way
interactions and could therefore not test this hypothesis.

\subsection{Simulation (80\%)}
Based on the above described movement model, we simulated 80'000 dispersers
moving 2'000 steps across the landscapes of the KAZA-TFCA. This simulation of
160 Mio. steps required five days of computation on a modern desktop machine.
The long simulation time was primarily caused by the massive extent considered
(ca. 1.8 Mio. km \textsuperscript{2} when including the buffer) and the large
number of dispersers simulated. Most connectivity studies are limited to much
smaller extents (e.g. \citealp{Kanagaraj.2013, Clark.2015, McClure.2016,
Abrahms.2017, Zeller.2020}) and will therefore require shorter simulation times.
We also believe that fewer simulated dispersers will often suffice, as the
relative traversal frequency by simulated individuals through randomly placed
checkpoints converged rather quickly. In any case, the exact number of required
simulations will vary depending on the structure of the landscape and the
dispersal ability of the focal species.

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
% Dispersal & Connectivity
Dispersal is an important process that allows species to avoid inbreeding,
colonize new habitats and reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and preservation of dispersal corridors has become a
task of extraordinary importance for conservation authorities.

% Quantifying Connectivity
Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their usefulness in reality. Individual-based dispersal simulations have
been proposed to address these shortcomings, yet due to the sheer amount of
non-trivial modeling decisions decisions, a unified and objective framework to
simulate dispersal is missing.

% Step Selection Functions
Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful to
investigate \textit{transient} movement behavior during dispersal and landscape
connectivity.

% Current Study
Here, we showcase the use of integrated step selection functions as simple,
individual-based, and spatially explicit movement model to simulate dispersal of
the endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions, thereby parametrizing a fully
mechanistic movement model describing how dispersing wild dogs move through the
available landscape. Based on this model, we simulate 80'000 dispersers moving
across the extent of the KAZA-TFCA and generate a set of three connectivity
maps, each focused on a different aspect of connectivity. Finally, we discuss
the benefits and pitfalls of such a simulation-based approach and highlight
potential improvements to be made in the future.

@@

% Create linenumbers
% \linenumbers

@@

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
analysis (LCPA; \citealp{Adriaensen.2003}) and circuit theory (CT;
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape to infer likely movement corridors. Despite their
intuitive nature and ease of use, both methods make rigorous assumptions about
animal movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be fulfilled by
migrating animals that typically move between a discrete set of habitats through
familiar landscapes. Dispersers, on the other hand, usually move over long
distances into unknown territory and are therefore less likely to be aware of
associated movement costs \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Another
issue of LCPs analysis concerns the fact that least-costly routes, by their very
nature, are only one pixel wide \citep{Pinto.2009}. This neglects the fact that
alternative routes with similar costs may exist and implies that the width of
inferred movement routes depends on the resolution of chosen covariate layers
and may not be biologically meaningful \citep{Diniz.2020}. Although some of
these deficiencies can be addressed using less stringent versions of the least
cost algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness remains.

% Issues with Circuit Theory
\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are rarely ever met. For
example, because CT only allows movements from a source cell to its four or
eight adjacent cells, it implicitly posits that animals exhibit a perceptual
range of a single pixel. Given that covariate layers are usually resolved with a
pixel size between 30 m x 30 m and 1 km x 1 km, this hardly ever renders the
true capability of animals to perceive the environment. Moreover, CT is built
around the assumption of a complete random walk \citep{Diniz.2020}, entailing
that directional biases cannot be rendered. Nevertheless, directionality is a
common characteristic in animal movement \citep{Bovet.1991, Schultz.2001},
especially in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}.

% Issues with Both Methods
\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal \citep{Diniz.2020}. Statements about the expected
duration required to traverse a certain corridor are therefore impossible.
Likewise, because movement is not modeled explicitly, interactions between
movement and habitat preferences of the focal species cannot be rendered.
Connectivity therefore merely arises in result to the landscape structure, which
is usually referred to as structural connectivity. While structural connectivity
yields insights in the \textit{potential} of the landscape to be traversed, it
does not enable to quantify the \textit{actual} gene flow through the area.
Consequently, a functional view on connectivity, which also renders the
behavioral response of the animal with respect to prevailing habitat conditions,
is often more desirable \citep{Tischendorf.2000, Baguette.2013}.

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis as proposed
    in this article. First, GPS data and habitat covariates have to be
    collected. The combined data is then analyzed in an integrated step
    selection model, which enables the parametrization of the focal species'
    habitat and movement kernels. The parametrized model is then treated as an
    individual-based movement model and used to simulate dispersal trajectories.
    Ultimately, simulated trajectories serve to produce a set of maps that are
    pertinent to landscape connectivity. This includes a heatmap, indicating the
    relative traversal frequency across the study area, a betweenness map,
    highlighting movement corridors and bottlenecks, and, finally, an
    inter-patch connectivity map, where the frequency of connections and their
    average duration can be depicted. \textcolor{red}{Photos: Whom to cite?
    Vectronics or Photographers?}}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Study Area (90\%)}
The study area was defined by a bounding box centered at -17\degree 13'9''S,
23\degree 56'4''E (\Cref{StudyArea}a) stretching over 1.3 Mio.
km\textsuperscript{2} and encompassed the entire KAZA-TFCA (\Cref{StudyArea}b).
The KAZA-TFCA represents the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. It covers a
total of 520'000 km\textsuperscript{2} and hosts diverse landscapes, ranging
from savanna to grassland and from dry to moist woodland habitats. In its center
lies the Okavango Delta, a dominant hydrogeographical feature and the world's
largest flood-pulsing inland delta. The wet season within the KAZA-TFCA lasts
from November to March and is out of phase with the flood in the Okavango Delta,
which peaks between July and August \citep{McNutt.1996, Wolski.2017}. Although
large portions within the KAZA-TFCA are designated national parks or other
protected areas, considerable human influence remains due to roads, agricultural
sites and settlements and villages that are distributed across the KAZA-TFCA's
landscape.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, black] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, black] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    encompassed parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green).}
    \label{StudyArea}
  \end{center}
\end{figure}

@@

We ran conditional logistic regression analysis in the r-package {\tt glmmTMB}
to estimate preferences of interest. To handle multiple individuals, we applied
the mixed effects technique developed by \citep{Muff.2020}, which allows to
effectively model random slopes. Thus, we treated animal IDs as random effect
and modeled random slopes for each covariate.

@@

\subsection{Convergence (80\%)}
To verify that the number of simulated individuals sufficed to ensure reliable
estimates of connectivity, we evaluated how the relative traversal frequency
across the landscape changed as we increased the number of simulated
individuals. For this, we distributed 1'000 rectangular ``checkpoints'', each
with an extent of 5 km x 5 km at random locations inside the main study area. We
then determined the relative traversal frequency by simulated trajectories
through each checkpoint for different numbers of simulations (1 to 50'000). To
assess variability in the relative traversal frequency, we repeatedly sampled
trajectories 100 times and computed the mean traversal frequency across
replicates, as well as the 95\% prediction-interval. We deemed that a checkpoint
converged as soon as the width of the prediction interval for the traversal
frequency across replicates dropped below a value of 0.01.

@@

@@

\subsection{Inter-Patch Connectivity (80\%)}
Results from the analysis of inter-patch connectivity are given in
\Cref{AreasReached}. The figure shows all realized links by simulated dispersers
between national parks and indicates the average duration a disperser had to
move to realize those links. It is again worth pointing out that the figure is
only intended as an example, as for clarity we only considered connectivity
between national parks (NPs), albeit plenty of links between other protected
areas exist. As can be seen from the number, thickness, and color of arrows,
inter-patch connectivity between NPs in Angola, Namibia, and Botswana is
comparably high and dispersal events between those areas short. In contrast, we
see that connections into the Kafue NP in Zambia require more steps and are
fewer in general. Similarly, there is a lack of connections into Zimbabwe's
Chizarira and Matusadona NP and the more distant Lower Zambezi and Mana Pools
NPs. In some cases, one can also detect imbalances between ingoing and outgoing
links, hinting at potential source-sink dynamics that occur due to asymmetries
in landscape permeability depending on the origin. For instance, while a large
portion of dispersers from the Chizaria NP in Zimbabwe manage to move into the
Hwange NP, there are comparably few dispersers that succeed in the opposite
direction.

@@

\section{Discussion}

% Short Summary
\subsection{Short Summary (90\%)}
We used ISSFs to analyse data of dispersing wild dogs and to parametrize a fully
mechanistic movement model describing how dispersers move through the available
landscape. We employed the parametrized model as an individual-based movement
model to simulate 80'000 dispersing wild dogs moving 2'000 steps across the
extent of the KAZA-TFCA, the world's largest transboundary conservation area.
Based on simulated dispersal trajectories, we derived three complementary maps,
each geared towards a better understanding of dispersal and landscape
connectivity. The set of maps included a heatmap, revealing frequently traversed
areas, a betweenness-map, delineating critical dispersal corridors, and a map of
inter-patch connectivity, indicating the presence or absence of functional links
between national parks as well as the average dispersal duration required to
realize those links. We thereby showcase that ISSFs offer a simple, yet powerful
framework to parametrize movement models and simulate dispersal to assess
landscape connectivity. Importantly, individual-based simulations from ISSFs
overcome several conceptual shortcomings inherent to more traditional
connectivity modeling techniques, such as least-cost path analysis and circuit
theory.

% Movement Model
\subsection{Movement Model (80 \%)}
Our movement model of dispersing wild dogs comprised a habitat kernel, a
movement kernel, and their interactions. Thus, the model rendered habitat and
movement preferences of dispersers, as well as how their movement preferences
were affected by habitat conditions. Parameter estimates from the habitat kernel
revealed that dispersers avoid water, prefer its proximity, avoid woodland,
prefer shrubs/grassland, and avoid areas dominated by humans. These results are
consistent with findings from previous studies on dispersing wild dogs
\citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019, Oneill.2020}, as well
as with a base model that we developed focussing on dispersing wild dogs'
habitat kernel \citep{Hofmann.2021}.

By expanding the base model by a proper movement kernel, we were able to model
several additional complexities inherent to dispersal. For instance, it is well
known that dispersers usually move with directional persistence
\citep{Cozzi.2018, Hofmann.2021} and that step lengths are typically correlated
with turning angles \citep{Morales.2004, Borger.2012}. That is, larger steps
usually coincide with smaller turning angles and vice versa. While such behavior
could be captured by jointly sampling turning angles and step lengths from
copula probability distributions \citep{Hodel.2021}, the ISSF framework allowed
us to model similar behavior directly using the movement model. Besides
accounting for directional persistence and correlations between step lengths and
turning angles, we also allowed for interactions that rendered the fact that
wild dogs mainly move during the darker morning and evening hours, whereas they
tend to rest during the remainder of the day.

By allowing interactions between habitat covariates and movement covariates, we
furthermore accounted for the fact that movement and habitat preferences are
interdependent. For example, the final model retained an interaction between
water cover and step length, showing that dispersers are more likely to realize
short steps (i.e. move slower) in areas covered by water and large steps in
areas located on dryland. Likewise, the parameter estimate for the interaction
between water cover and turning angles revealed that dispersers move less
directional across water bodies than across dryland. We believe that this is
owed to the fact that wild dogs wade or swim when traversing waterbodies, thus
resulting in slower, more tortuous movements. Besides this, our model also
suggested that dipsersers preferably realize shorter steps when moving through
woodland, but larger steps when moving across shrubs/grassland. This can likely
be linked to wild dogs' resting behavior, as wild dogs usually use open areas to
quickly move over long distances \citep{Abrahms.2017} but seek shade and
protection below the woodland canopy when resting \citep{Creel.2002}.

\subsection{Simulation (80\%)}
Based on the above described movement model, we simulated 80'000 dispersers
moving 2'000 steps across the landscapes of the KAZA-TFCA. On a modern desktop
machine, this simulation required five days of computation. The long simulation
time was primarily caused by the massive extent considered (ca. 1.8 Mio. km
\textsuperscript{2} when including the buffer) and the large number of
dispersers simulated. Most connectivity studies are limited to much smaller
extents (e.g. \citealp{Kanagaraj.2013, Clark.2015, McClure.2016, Abrahms.2017,
Zeller.2020}) and will therefore achieve faster simulation times. We also
believe that fewer simulated dispersers will often suffice, as the relative
traversal frequency by simulated individuals through randomly placed checkpoints
in the study area converged already after 10'500 simulated individuals in our
case. The required number of simulated individuals will, however, vary depending
on the structure of the landscape and the dispersal ability of the focal
species.
